### 2016-07-05 introduction.tex
C --> B. KC's suggestions incorporated.

Recent results on reasoning about application invariants in the
presence of weak consistency~\cite{burckhardt14, redblueosdi,
redblueatc, ecinec, gotsmanpopl16} address broadly related concerns.
Weak consistency is a phenomenon that usually manifests on replicated
data stores, where operations are concurrently executed against
different replicas, resulting in an order of execution inconsistent
with their serial order. The operations, nonetheless, are atomic and
fully isolated, and each operation is required to locally preserve
application invariants. In contrast, weak isolation is a property of
transactions, which are sets of atomic operations. Weak isolation
manifests when successive atomic operations in a transaction witness
different contemporary states (or different replicas) of the database
which, although consistent individually, may not be obviously
reconciled into a consistent global view. Furthermore, operations may
temporarily break invariants as long as transaction as a whole
preserves them.  Intuitively, the latter problem reduces to the former
if all transactions contain a single operation. Furthermore, weak
isolation can exist independent of weak consistency, as evident from
the presence of weakly-isolated transactions on sequentially
consistent relational databases; such systems existed long before the
advent of weakly consistent NoSQL stores.  Thus, frameworks for
reasoning about weak isolation will necessarily have to generalize the
reasoning frameworks developed for weak consistency in new and
important ways.

@@

\noindent Our results provide the first (to the best of our knowledge)
mechanism that precisely and uniformly relates high-level program
invariants to low-level weak isolation and store consistency
guarantees, thereby allowing weakly-isolated transactions to enjoy the
same rigorous reasoning capabilities as their strongly-isolated
(serializable) counterparts.

>>>>

Recent results on reasoning about application invariants in the
presence of weak consistency~\cite{burckhardt14, redblueosdi,
redblueatc, ecinec, gotsmanpopl16} address broadly related concerns.
Weak consistency is a phenomenon that usually manifests on replicated
data stores, where operations are concurrently executed against
different replicas, resulting in an order of execution inconsistent
with their serial order. The operations, nonetheless, are atomic and
fully isolated, and each operation is required to locally preserve
application invariants. In contrast, weak isolation is a property of
transactions, which are sets of atomic operations. Weak isolation
manifests when successive atomic operations in a transaction witness
different contemporary states of the database (or different replicas
of the replicated store) which, although consistent individually, may
not be obviously reconciled into a consistent global view.
Intuitively, the latter problem reduces to the former if all
transactions contain a single operation. Furthermore, weak isolation
can exist independent of weak consistency, as evident from the
presence of weakly-isolated transactions on conventional RDBS. With an
added complexity of replication, richer mechanisms are needed to
reason about weak isolation in tandem with weak consistency. Thus,
frameworks for reasoning about weak isolation will necessarily have to
generalize the reasoning frameworks developed for weak consistency in
new and important ways.

@@

\noindent Our results provide the first (to the best of our knowledge)
mechanism that precisely and uniformly relates high-level program
invariants to low-level weak isolation and weak consistency
guarantees, thereby allowing weakly-isolated transactions to enjoy the
same rigorous reasoning capabilities as their strongly-isolated
(serializable) counterparts.

### 2016-07-05 motivation.tex
C --> B. KC's suggestions incorporated.

In this section, we motivate our ideas via an example written in
\txnimp - a C-like imperative language equipped with a \C{txn} lexical
block defining a transaction scope.  For exposition convenience, each
\C{txn} block is associated with a single-quoted string in angle
braces that uniquely identifies the transaction. We use Hoare triple
notation to annotate programs with pre- and post- conditions.

\begin{figure}
\centering
$\{\texttt{C}=\texttt{k} \conj \texttt{k}\ge\texttt{a1+a2}\}$
\begin{tabular}{l||l}
\begin{txnimpcode}
  txn$\langle$'Wd1'$\rangle${
    if (C $\ge$ a1) {
      C := C - a1
    }
  }
\end{txnimpcode}
&
\begin{txnimpcode}
  txn$\langle$'Wd2'$\rangle${
    if (C $\ge$ a2) {
      C := C - a2
    }
  }
\end{txnimpcode}
\\
\end{tabular}
$\{\texttt{C}=\texttt{k-a1-a2}\}$
\caption{\small Concurrent withdraw transactions}
\label{fig:motiv-eg-1}
\vspace*{-10pt}
\end{figure}

Consider an implementation of a banking application that admits
concurrent withdraw transactions on a checking account (\C{C}), as
shown in Fig.~\ref{fig:motiv-eg-1}. If the initial balance (\C{k}) in
the account is enough to perform both withdraws, then the final
balance, after both transactions commit, is expected to reflect the
effects of both withdraws. The pre- and post-conditions in
Fig.~\ref{fig:motiv-eg-1} reflect this expectation. Indeed, invariants
are guaranteed to hold if both withdraw transactions are serialized,
making \iso{Serializable} isolation ({\sc ser}) level a sufficent
condition to preserve invaraints. But, is {\sc ser} necessary?

As an alternative, consider the execution of this transaction under
a \emph{Read Committed} ({\sc rc}) isolation level, which is weaker than
{\sc ser}.\footnote{{\sc rc} is in fact the default isolation level in
Postgres 9.5 and Oracle 11g databases.} An {\sc rc} transaction is
isolated from the writes of uncommitted transactions, and is therefore
free from \emph{dirty reads}~\cite{berenson} that read uncommitted
data. In the current example, {\sc rc} isolation admits the two
executions shown in Fig.~\ref{fig:rc-ex} on a strongly consistent
({\sc sc}) store, such as a relational database.

@@

The figure depicts an execution as a series of read, write, and commit
operations.  The subscript of an operation indicates the transaction
that executes it. In the first execution, transaction \C{Wd1} reads
the current balance (\C{k}) and writes the new balance (\C{k-a1}), but
before it commits, transaction \C{Wd2} executes and commits, writing
the new balance (\C{k-a2}). {\sc rc} isolation prevent \C{Wd2} from
witnessing the uncommitted writes of transaction \C{Wd1}.
Subsequently committing \C{Wd1} leads to the loss of \C{Wd2}'s updates
(the so-called \emph{lost update anomaly}~\cite{berenson}), resulting
in an incorrect balance of \C{k-a1}. The second execution describes a
similar scenario with \C{Wd1} and \C{Wd2} exchanging their roles.
Clearly, {\sc rc} is an excessively weak isolation level for this
program because it loses the updates of one transaction, resulting in
the violation of the post-condition.  A stronger isolation level that
prevents lost updates is required. \iso{Snapshot Isolation} ({\sc
si})~\cite{berenson} fits this requirement; {\sc si} effectively
serializes transactions that update a shared data object by aborting
and re-executing a transaction if write-write conflicts are detected
during its commit.\footnote{{\sc si} however does not serialize
transactions in the absence of write-write conflicts.} Since {\sc si},
unlike {\sc ser}, does not rely on expensive lock-based concurrency
control, it is also more efficient, making it appropriate for both
\C{Wd} transactions.

@@

\begin{figure}
\begin{smathpar}
\begin{array}{lcl}
\psi_{RC} & \defeq & \forall T_1,T_2,\eta_1,\eta_2,.\; \txn(\eta_1) = T_1 
  \conj \txn(\eta_2) = T_2 \\
  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
                      {T_2 \wrstoar \C{X}})\\
  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
\end{array}
\end{smathpar}
\caption{\small Visibility and interference properties for different weak isolation levels can be
  captured by different instantiations of their happens-before relation.}
\label{fig:interference-ex}
\end{figure}

@@

For the rely-guarantee approach to be useful, it has to thus constrain
allowed interference in accordance with the chosen level of isolation
associated with a transaction. The first observation we make is that
we can constrain interference by constraining the definition of the
\emph{happens-before} ($\hbZ$) relation, which ultimately dictates
what actions become visible to a transaction, or its constituents, and
when. To do so, we axiomatize the $\hbZ$ relation to capture the
interference characteristics of different isolation levels as their
specifications. Two such specifications are shown\footnote{Note that
specifications presented here are solely for illustration. The actual
specifications described in \S\ref{sec:ansi-isolation} are more
nuanced.} in Fig.~\ref{fig:interference-ex}. The \iso{Read Committed}
specification ($\psi_{RC}$) allows an operation $\eta_1$ of $T_1$ to
happen-before an operation $\eta_2$ of $T_2$ only if every operation
in $T_1$ (including its commit) happens-before $\eta_2$ (we let $T_1
\hboar \eta_2$ denote this). The specification does not require $T_1$
to execute and commit before all operations of $T_2$, thus allowing it
to interfere in $T_2$. The \iso{Snapshot Isolation} specification
($\psi_{SI}$) requires two transactions that write to the same
variable to be related by happens-before.  This effectively prohibits
interference due to actions in $T_1$ being interleaved in $T_2$ (or,
vise versa).

Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ from \C{Wd2}
to \C{Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. All arrows are
legal under $\psi_{RC}$ because in no case does an operation $\eta_1$
from \C{Wd1} happen-before $\eta_2$ of \C{Wd2} without the commit of
\C{Wd1} also happening before $\eta_2$. $\psi_{SI}$ however disallows
all dashed arrows, because dashed arrows establish $\hbZ$ edges
between actions in \C{Wd2} and only a subset of the operations in
\C{Wd1}.  Since the result of an operation (\eg a read) depends on
what happens before that operation, the structure of the $\hbZ$
relation ultimately dictates the final result of the program. The
solid arrow and the first (from left) dashed arrow in
Fig.~\ref{fig:motiv-eg-1-hb} denote $\hbZ$ relationships that do not
affect the value of \C{C} in a way that causes the program to violate
its post-condition. The last two dashed arrows denote $\hbZ$
relationships that lead to the violation of the post-condition. The
task of the reasoning framework is to determine if all $\hbZ$
relationships allowed by an isolation level lead to the satisfaction
of the post-condition.

@@

The specification of an isolation level encodes its interference
characteristics as constraints over the $\hbZ$ relation. However, for
this to be useful in reasoning about programs, the rely-guarantee
framework should be able to use the specification to determine if an
interference is valid or not, allowing the programmer to only
focus on the former. Our second observation is that this is possible
if the reasoning framework adequately tracks $\hbZ$ at each program
point, while preserving $\hbZ$ constraints as invariants between
program points. An interference that leads to the violation of $\hbZ$
constraints (i.e., an invalid interference) is thus automatically
prohibited. For instance, consider the program point after the write
to \C{C} in \C{Wd1}. The expected invariant ($\phi$) at that program
point is shown below ($\committed$ stands for ``committed''):

\begin{smathpar}
\begin{array}{c}
 \neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{C}  \conj 
    (\neg\committed(\C{Wd2}) \Rightarrow \C{C = k-a1}) \\
    \conj (\committed(\C{Wd2})
                \Rightarrow \C{C = k-a1-a2})
\end{array}
\end{smathpar}

\noindent $\phi$ asserts that \C{Wd1} is not yet committed, and that
it wrote to \C{C}, and the value of \C{C} is either \C{k-a1-a2} or
\C{k-a1} depending on whether or not \C{Wd2} is committed. If $\phi$
remains invariant until \C{Wd1} commits, then the post-condition (\C{C
= k-a1-a2}) can be established easily. However, an interference from
\C{Wd2} at this stage (captured by the last dashed arrow in
Fig.~\ref{fig:motiv-eg-1-hb}) may violate the invariant by writing
$\C{k-a1}$ to \C{C} and committing \C{Wd2}, thus leading to
$\committed(\C{Wd2}) \Rightarrow \C{C=k-a2}$. Fortunately,
\iso{Snapshot Isolation} prevents this interference, and this can be
shown by demonstrating that an interference from \C{Wd2} starting from
an execution state that satisfies $\psi_{SI} \wedge \phi$ leads to an
execution state where neither $\C{Wd2} \hboar \C{Wd1}$ nor $\C{Wd1}
\hboar \C{Wd2}$ holds ($\C{Wd2} \hboar \C{Wd1}$ does not hold because
\C{Wd1}'s write to \C{C} clearly happened before \C{Wd2}'s commit, and
$\C{Wd1} \hboar \C{Wd2}$ does not hold because \C{Wd1} has not yet
committed, while \C{Wd2} has already begun). Since \C{Wd2} also writes
to \C{C}, this violates the $\psi_{SI}$ constraint which we assume to
be an invariant. A proof that the post-condition holds now follows
from the contradiction. It is informative to note that if the
invariant is $\psi_{RC}$ instead of $\psi_{SI}$, we cannot derive a
contradiction and we cannot rule out the interference, which
(rightfully) causes  the proof to fail.

We have thus far assumed a strongly consistent ({\sc sc}) store that,
in the absence of transactions with special isolation requirements,
makes the effects of any operation immediately visible to all
subsequent operations. The natural behavior of {\sc sc} to totally
order all operations w.r.t. the $\hbZ$ relation could be in conflict
with the constraints imposed by weak isolation. To meet the
requirements of weaker isolation levels, say {\sc rc}, the store has
to adapt itself to \emph{hide} the effects of concurrent transactions
until they are committed. Once committed though, effects need to be
immediately visible to subsequent operations. This semantics can be
built into the reasoning framework, leading to a proof system
tailor-made for such stores. However, for the reasoning framework to
be truly useful, it has to be parametric over different store
consistency semantics, supporting stores that are weaker than {\sc sc}
(e.g., causally or eventually consistent), and should be able to
reconcile conflicts between consistency and isolation constraints.  We
demonstrate how our reasoning framework makes this possible in the
following sections.

>>>>

In this section, we motivate our ideas via an example written in
\txnimp - a C-like imperative language equipped with a \C{txn} lexical
block defining a transaction scope.  For expositional convenience,
each \C{txn} block is associated with a single-quoted string in angle
braces that uniquely identifies the transaction. We use Hoare triple
notation to annotate programs with pre- and post- conditions.

\begin{figure}
\centering
$\{\texttt{B}=\texttt{k} \conj \texttt{k}\ge\texttt{a1+a2}\}$
\begin{tabular}{l||l}
\begin{txnimpcode}
  txn$\langle$'Wd1'$\rangle${
    if (B $\ge$ a1) {
      B := B - a1
    }
  }
\end{txnimpcode}
&
\begin{txnimpcode}
  txn$\langle$'Wd2'$\rangle${
    if (B $\ge$ a2) {
      B := B - a2
    }
  }
\end{txnimpcode}
\\
\end{tabular}
$\{\texttt{B}=\texttt{k-a1-a2}\}$
\caption{\small Concurrent withdraw transactions}
\label{fig:motiv-eg-1}
\vspace*{-10pt}
\end{figure}

Consider an implementation of a banking application that admits
concurrent withdraw transactions on an account balance (\C{B}), as
shown in Fig.~\ref{fig:motiv-eg-1}. If the initial balance (\C{k}) in
the account is enough to perform both withdraws, then the final
balance, after both transactions commit, is expected to reflect the
effects of both withdraws. The pre- and post-conditions in
Fig.~\ref{fig:motiv-eg-1} reflect this expectation. Indeed, invariants
are guaranteed to hold if both withdraw transactions are serialized,
making \iso{Serializable} isolation ({\sc ser}) level a sufficent
condition to preserve invaraints. But, is {\sc ser} necessary?

As an alternative, consider the execution of this transaction under a
\emph{Read Committed} ({\sc rc}) isolation level, which is weaker than
{\sc ser}.\footnote{{\sc rc} is in fact the default isolation level in
Postgres 9.5 and Oracle 11g databases.} An {\sc rc} transaction is
isolated from the writes of uncommitted transactions, and is therefore
free from \emph{dirty reads}~\cite{berenson} that read uncommitted
data. In the current example, {\sc rc} isolation admits the two
executions shown in Fig.~\ref{fig:rc-ex} on a strongly consistent
({\sc sc}) store, such as a conventional RDBMS.

@@

The figure depicts an execution as a series of read, write, and commit
operations. The subscript of an operation indicates the transaction
that executes it (Both transactions read \C{B} twice, hence two read
operations). In the execution on left, transaction \C{Wd1} reads the
current balance (\C{k}) and writes the new balance (\C{k-a1}), but
before it commits, transaction \C{Wd2} executes and commits, writing
the new balance (\C{k-a2}). {\sc rc} isolation prevent \C{Wd2} from
witnessing the uncommitted writes of transaction \C{Wd1}.
Subsequently committing \C{Wd1} leads to the loss of \C{Wd2}'s updates
(the so-called \emph{lost update anomaly}~\cite{berenson}), resulting
in an incorrect balance of \C{k-a1}. The execution on right describes
a similar scenario with \C{Wd1} and \C{Wd2} exchanging their roles.
Clearly, {\sc rc} is an excessively weak isolation level for this
program because it loses the updates of one transaction, resulting in
the violation of the post-condition.  A stronger isolation level that
prevents lost updates is required. \iso{Snapshot Isolation} ({\sc
si})~\cite{berenson} fits this requirement; {\sc si} effectively
serializes transactions that update a shared data object by aborting
and re-executing a transaction if write-write conflicts are detected
during its commit.\footnote{{\sc si} however does not serialize
transactions in the absence of write-write conflicts.} Since {\sc si},
unlike {\sc ser}, does not need expensive mechanisms such as
lock-based concurrency control or runtime monitoring, it is also more
efficient, making it appropriate for both withdraw transactions.

@@

\begin{figure}
\begin{smathpar}
\begin{array}{lcl}
\psi_{RC} & \defeq & \forall T_1,T_2,\eta_1,\eta_2.\; \txn(\eta_1) = T_1 
  \conj \txn(\eta_2) = T_2 \\
  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
                      {T_2 \wrstoar \C{X}})\\
  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
\end{array}
\end{smathpar}
\caption{\small Visibility and interference properties for different weak isolation levels can be
  captured by different instantiations of their happens-before relation.}
\label{fig:interference-ex}
\end{figure}

@@

For the rely-guarantee approach to be useful, it has to thus constrain
allowed interference in accordance with the chosen level of isolation
associated with a transaction. The first observation we make is that
we can constrain interference by constraining the definition of the
\emph{happens-before} ($\hbZ$) relation, which ultimately dictates
what actions become visible to a transaction, or its constituents, and
when. To do so, we axiomatize the $\hbZ$ relation to capture the
interference characteristics of different isolation levels as their
specifications. Two such specifications are shown\footnote{Note that
specifications presented here are solely for illustration. The actual
specifications described in \S\ref{sec:ansi-isolation} are more
nuanced.} in Fig.~\ref{fig:interference-ex}. The \iso{Read Committed}
specification ($\psi_{RC}$) allows an operation $\eta_1$ of $T_1$ to
happen-before an operation $\eta_2$ of $T_2$ only if every operation
in $T_1$ (including its commit) happens-before $\eta_2$ (we let $T_1
\hboar \eta_2$ denote this). The specification does not require $T_1$
to execute and commit before all operations of $T_2$, thus allowing it
to interfere in $T_2$. The \iso{Snapshot Isolation} specification
($\psi_{SI}$) requires two transactions that write to the same
variable (denoted $\wrstoar$)to be related by happens-before.  This
effectively prohibits interference due to actions in $T_1$ being
interleaved in $T_2$ (or, vise versa).

Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ from \C{Wd2}
to \C{Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. All arrows are
legal under $\psi_{RC}$ because in no case does an operation $\eta_1$
from \C{Wd1} happen-before $\eta_2$ of \C{Wd2} without the commit of
\C{Wd1} also happening before $\eta_2$. $\psi_{SI}$ however disallows
all dashed arrows, because dashed arrows establish $\hbZ$ edges
between actions in \C{Wd2} and only a subset of the operations in
\C{Wd1}.  Since the result of an operation (\eg a read) depends on
what happens before that operation, the structure of the $\hbZ$
relation ultimately dictates the final result of the program. The
solid arrow and the first (from left) dashed arrow in
Fig.~\ref{fig:motiv-eg-1-hb} denote $\hbZ$ relationships that do not
affect the value of \C{B} in a way that causes the program to violate
its post-condition. The last two dashed arrows denote $\hbZ$
relationships that lead to the violation of the post-condition. The
task of the reasoning framework is to determine if all $\hbZ$
relationships allowed by an isolation level lead to the satisfaction
of the post-condition.

@@

The specification of an isolation level encodes its interference
characteristics as constraints over the $\hbZ$ relation. However, for
this to be useful in reasoning about programs, the rely-guarantee
framework should be able to use the specification to determine if an
interference is valid or not, allowing the programmer to only
focus on the former. Our second observation is that this is possible
if the reasoning framework adequately tracks $\hbZ$ at each program
point, while preserving $\hbZ$ constraints as invariants between
program points. An interference that leads to the violation of $\hbZ$
constraints (i.e., an invalid interference) is thus automatically
prohibited. For instance, consider the program point after the write
to \C{B} in \C{Wd1}. The expected invariant ($\phi$) at that program
point is shown below ($\committed$ stands for ``committed''):

\begin{smathpar}
\begin{array}{c}
 \neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{B}  \conj 
    (\neg\committed(\C{Wd2}) \Rightarrow \C{B = k-a1}) \\
    \conj (\committed(\C{Wd2})
                \Rightarrow \C{B = k-a1-a2})
\end{array}
\end{smathpar}

\noindent $\phi$ asserts that \C{Wd1} is not yet committed, and that
it wrote to \C{B}, and the value of \C{B} is either \C{k-a1-a2} or
\C{k-a1} depending on whether or not \C{Wd2} is committed. If $\phi$
remains invariant until \C{Wd1} commits, then the post-condition (\C{B
= k-a1-a2}) can be established easily. However, an interference from
\C{Wd2} at this stage (captured by the last dashed arrow in
Fig.~\ref{fig:motiv-eg-1-hb}) may violate the invariant by writing
$\C{k-a1}$ to \C{B} and committing \C{Wd2}, thus leading to
$\committed(\C{Wd2}) \Rightarrow \C{B=k-a2}$. Fortunately,
\iso{Snapshot Isolation} prevents this interference, and this can be
shown by demonstrating that an interference from \C{Wd2} starting from
an execution state that satisfies $\psi_{SI} \wedge \phi$ leads to an
execution state where neither $\C{Wd2} \hboar \C{Wd1}$ nor $\C{Wd1}
\hboar \C{Wd2}$ holds; $\C{Wd2} \hboar \C{Wd1}$ does not hold because
\C{Wd1}'s write to \C{B} clearly happened before \C{Wd2}'s commit, and
$\C{Wd1} \hboar \C{Wd2}$ does not hold because \C{Wd1} has not yet
committed, while \C{Wd2} has already begun. Since \C{Wd2} also writes
to \C{B}, this violates the $\psi_{SI}$ constraint which we assume to
be an invariant. A proof that the post-condition holds now follows
from the contradiction. It is informative to note that if the
invariant is $\psi_{RC}$ instead of $\psi_{SI}$, we cannot derive a
contradiction and we cannot rule out the interference, which
(rightfully) causes  the proof to fail.

We have thus far assumed a strongly consistent ({\sc sc}) store that,
in the absence of transactions with special isolation requirements,
makes the effects of any operation immediately visible to all
subsequent operations. The natural behavior of {\sc sc} to totally
order all operations w.r.t. the $\hbZ$ relation could be in conflict
with the constraints imposed by weak isolation. To meet the
requirements of weaker isolation levels, say {\sc rc}, the store has
to adapt itself to \emph{hide} the effects of concurrent transactions
until they are committed. Once committed though, effects need to be
immediately visible to subsequent operations. This semantics can be
built into the reasoning framework, leading to a proof system
tailor-made for such stores. However, for the reasoning framework to
be truly useful, it has to be parametric over different consistency
semantics, supporting stores that are weaker than {\sc sc} (e.g.,
causally or eventually consistent), and should be able to reconcile
conflicts between consistency and isolation constraints.  We
demonstrate how our reasoning framework makes this possible in the
following sections.

### 2016-07-05 reasoning.tex
C --> B. KC's suggestions incorporated.

As mentioned earlier, our proof system is a rely-guarantee framework
that admits compositional reasoning by abstracting away environmental
interference as a rely relation. A conspicuous difference between a
standard development of rely-guarantee and ours is that, while the
former reasons in terms of program states (variable to value
bindings), we reason in terms of executions as captured by their
traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
relations relate executions (i.e., $R,\,Q\,\subseteq\,\E\times\E$),
and our pre- ($P$) and post- ($Q$) conditions are assertions over
executions (i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development
also facilitates state-based reasoning via the interpretation function
($\interp{\cdot}$) introduced in \S\ref{sec:syntax}. For example, if a
\emph{bi-state} rely relation relates every pair of states $\sigma$
and $\sigma'$ such that $\sigma'(X) \ge \sigma(X) \ge 0$, the
corresponding \emph{bi-execution} rely relation relates every pair of
executions $\E$ and $\E'$ such that $\interp{\E}(X) \ge
\interp{\E'}(X) \ge 0$.  Assertions on states are also written
similarly. For instance, consider the post-condition
$\C{C\,=\,k-a1-a2}$ of the program in Fig.~\ref{fig:motiv-eg-1}. The
corresponding assertion on the post-state
($\lambda\sigma.~\sigma(\C{C}) \,=\, \C{k-a1-a2}$) asserts that in all
states resulting from executing the program, the value of \C{C} is
\C{k-a1-a2}. The equivalent execution-based assertion
($\lambda\E.~\interp{\E}(\C{C})\,=\,\C{k-a1-a2}$) asserts that in all
executions of the program, the value of \C{C} is \C{k-a1-a2}.
However, having access to an execution facilitates assertions that go
beyond the state. An example is an invariant of \C{Wd1} described
abstractly in \S\ref{sec:motivation}, and reified as an
execution-based assertion below:

\begin{smathpar}
\begin{array}{l}
  \lambda\E.~ \neg(\underE{\committed(\C{Wd1}}) \conj 
      \underE{\C{Wd1} \wrstoar \C{C}}\\
  \hspace*{0.2in}\conj(\neg(\underE{\committed(\C{Wd2})}) \Rightarrow 
      \C{\interp{\E}(C) = k-a1}) \\
   \hspace*{0.2in}\conj (\underE{\committed(\C{Wd2})} \Rightarrow 
      \C{\interp{\E}(C) = k-a1-a2})
\end{array}
\end{smathpar}
% Where, $\committed$ is defined as:
% \begin{smathpar}
% \begin{array}{lcl}
% \underE{\committed(T_i)} & \Leftrightarrow &
%   \exists\eta.~\underE{\eta\in T_i} \conj \kind(\eta) = \C{COMMIT}\\
% \end{array}
% \end{smathpar}

\noindent One of the conjuncts asserts that \C{Wd1} has written to \C{C} - a
fact which cannot be deduced solely from the value of \C{C} (esp. in the
presence of interference), but can be expressed as a proposition over
the execution trace.

>>>>

As mentioned earlier, our proof system is a rely-guarantee framework
that admits compositional reasoning by abstracting away environmental
interference as a rely relation. A conspicuous difference between a
standard development of rely-guarantee and ours is that, while the
former reasons in terms of program states (variable to value
bindings), we reason in terms of executions as captured by their
traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
relations relate executions (i.e., $R,\,Q\,\subseteq\,\E\times\E$),
and our pre- ($P$) and post- ($Q$) conditions are assertions over
executions (i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development
also facilitates state-based reasoning via the interpretation function
($\interp{\cdot}$) introduced in \S\ref{sec:syntax}. For example, if a
\emph{bi-state} rely relation relates every pair of states $\sigma$
and $\sigma'$ such that $\sigma'(X) \ge \sigma(X) \ge 0$, the
corresponding \emph{bi-execution} rely relation relates every pair of
executions $\E$ and $\E'$ such that $\interp{\E}(X) \ge
\interp{\E'}(X) \ge 0$.  Assertions on states are also written
similarly. For instance, consider the post-condition
$\C{B\,=\,k-a1-a2}$ of the program in Fig.~\ref{fig:motiv-eg-1}. The
corresponding assertion on the post-state
($\lambda\sigma.~\sigma(\C{B}) \,=\, \C{k-a1-a2}$) asserts that in all
states resulting from executing the program, the value of \C{B} is
\C{k-a1-a2}. The equivalent execution-based assertion
($\lambda\E.~\interp{\E}(\C{B})\,=\,\C{k-a1-a2}$) asserts that in all
executions of the program, the value of \C{B} is \C{k-a1-a2}.
However, having access to an execution facilitates assertions that go
beyond the state. An example is an invariant of \C{Wd1} described
abstractly in \S\ref{sec:motivation}, and reified as an
execution-based assertion below:

\begin{smathpar}
\begin{array}{l}
  \lambda\E.~ \neg(\underE{\committed(\C{Wd1}}) \conj 
      \underE{\C{Wd1} \wrstoar \C{B}}\\
  \hspace*{0.2in}\conj(\neg(\underE{\committed(\C{Wd2})}) \Rightarrow 
      \C{\interp{\E}(B) = k-a1}) \\
   \hspace*{0.2in}\conj (\underE{\committed(\C{Wd2})} \Rightarrow 
      \C{\interp{\E}(B) = k-a1-a2})
\end{array}
\end{smathpar}
% Where, $\committed$ is defined as:
% \begin{smathpar}
% \begin{array}{lcl}
% \underE{\committed(T_i)} & \Leftrightarrow &
%   \exists\eta.~\underE{\eta\in T_i} \conj \kind(\eta) = \C{COMMIT}\\
% \end{array}
% \end{smathpar}

\noindent One of the conjuncts asserts that \C{Wd1} has written to \C{B} - a
fact which cannot be deduced solely from the value of \C{B} (esp. in the
presence of interference), but can be expressed as a proposition over
the execution trace.
